---
title: "Lab 3: July 15"
author: "POLS 3316, Summer 2022, University of Houston, Instructor: Tom Hanna"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## Part 1: Scalars, Vectors, Matrices and Arrays

### What is a matrix?

A *matrix* is a rectangular array or table of numbers, symbols, or expressions, arranged in rows and columns, used to represent some mathematical object or property. 

Practical examples of *matrices*? Spreadsheets and tables

#### Matrixes in R

In R, every *dataframe* or *table* or similar *object* is a matrix.

For an example of building a matrix in R, we need to look at scalars and vectors. 


### What is a scalar?

A scalar is a single number. A scalar is matrix with dimension 1x1, 1 column by 1 row. A scalar is also a vector with only 1 element.

#### Scalars in R

To create a scalar in R, we can either assign a null value and add the value later, or we can assign a value when we creat it:

```{r}
a <- c()
cat('a is a null scalar ')
a

b <- 5
cat('b is a scalar with value 5')
b

a <- 1
cat('Now I have given a the value 1')
a
```



### What is a vector?

A vector is a list of numbers. It is either a single column or a single row. It is a matrix with dimension 1 x n . 

#### Vectors in R?

To create a vector in R, we can assign a series of values either by just listing them or through a variety of functions including operations on other vectors:

```{r}

C <- c(1,2,4,5,6,7,8,9,10)
cat('C is a vector ')
C

D <- C*b
cat('D is a vector created by multiplying b * C ')
D

```

```{r}

E <- seq(-5, 5, by = .5)
cat('E is a sequence of numbers from -5 to 5, by 0.5s ')
E

```

```{r}

F <- dnorm(E, mean = 2.5, sd = 0.5)
cat('F is a function of E and the normal distribution ')
F


```

## Things to know

- Multiplying anything times a scalar is simply multiplying every element by that scalar. It's almost always feasible. 
- To do an operation on two vectors they must be the same length
- Multiplying matrices is a complex version of multipying vectors. - If you see errors related to matrix/vector size, it usually means you are multiplying two matrixes who aren't a match in column and row size. If this happens, you have two options:

https://www.mathsisfun.com/algebra/matrix-multiplying.html

or 

**ask for help from someone with a good grasp of basic matrix algebra**

## Back to matrices in R

We built a scalar object and a vector object. How do we build a matrix object?

Let's use two of the vectors we built, C and D:

```{r}
data_cd <- data.frame(C,D)
head(data_cd)

```




## Part 2: Measures of dispersion in R          

We're going back to the arrest data. We can use the R functions *var()* to get the **sample** *variance* and *sd()* to get the **sample** *standard deviation*. The numbers we figured were the **population** numbers. The difference in the formulas is that for the *variance*, the population version divides by the number of observations *n* and the sample version divides by *n-1*. We'll discuss the theoretical difference in detail this week, but for now the difference is this: In most cases we are only working with a sample of data and we are using it to estimate the full "population" or "universe" of data. The slight difference in formula is to account for this.  

For now, let's start by looking at the Arrest Data *sample variance* and *sample standard deviation.*


To get **sample variance** use the R function **var().**


```{r}
arrests_data <- USArrests[2:3]

urban_pop_variance <- var(arrests_data$UrbanPop)
cat('Urban Pop Percent Variance')
urban_pop_variance

assaults_variance <- var(arrests_data$Assault)
cat('Assaults Variance')
assaults_variance

mean(arrests_data$Assault)

```



To get **sample standard deviation** either take the **square root of sample variance** or use the R function **sd()**.

```{r}

cat('Standard deviations')
urban_pop_sd1 <- sd(arrests_data$UrbanPop)
urban_pop_sd1

urban_pop_sd2 <- sqrt(var(arrests_data$UrbanPop))
urban_pop_sd2

```


To get **population variance** multiply the R function var() times (n-1)/n:  var(x) * (n-1)/n

```{r}

urban_pop_popvar <- var(arrests_data$UrbanPop) * (50-1)/50
cat('Urban Pop population variance:')
urban_pop_popvar

assaults_popvar <- var(arrests_data$Assault) *(50-1)/50
cat('Assaults population variance:')
assaults_popvar

```


To get **population standard deviation**, the only option is to compute the population variance and takes its square root: **sqrt(populationvariance)** *You can't get it by multiplying the R function at least easily.*

```{r}

sqrt(urban_pop_popvar)
sqrt(assaults_popvar)

```





