---
title: 'Lab 6: July 20'
author: 'POLS 3316, Summer 2022, University of Houston, Instructor: Tom Hanna'
output: html_notebook
---

##Probabilities with R frequency distribution functions

In this lab, we use the distribution functions from R for calculating probabilities.

1. Create a variable x with values (-2,-1,0,1,2)

```{r}

x <- c(-2,-1,0,1,2)

```

2. Calculate the value of the cumulative distribution function at (or the probability to the left of) a given value of x. Example with minus -2: 

```{r}
#pnorm gives the probability to the left of (x)
pnorm(-2)

```
3. Create a vector that contains the pnorm values for all values of x.

```{r}

pnorm_x <- pnorm(x)

```

4. Now create a variable “y” with values: (0,1,2,5,8,10,15,20)

```{r}

y <- c(0,1,2,5,8,10,15,20)

```


5. Using a binomial distribution, calculate the value of the cumulative distribution function at a given value of y. Example: 0

```{r}

pbinom(0,size=20,prob=.2)

```

6. Now create a vector with the binomial probabilities of all y values. 

```{r}

pbin_y <- pbinom(y,size=20,prob=.2)

```

7. Repeat Question 6 but using the Poisson and Uniform Distributions - commands are *ppois()* and *puni* - create vectors named vector_y_pois and vector_y_uni

```{r}



```

## Calculating probabilities with the functions: 

8. Probability that a normal random variable with mean 22 and variance 25 lies between 16.2 and 27.5. 


```{r}
#First, we need the standard deviation
sqrt(25)

#we use the pnorm function - the probability distribution function #for the normal distribution
#The first argument is the vector (or scalar) we want to compute
#The second argument is the mean of the distribution (22)
#The third argument is the standard deviation 
#We subtract one result from the other
answer8 <- pnorm(27.5,22,sd=5) - pnorm(16.2,22,sd=5)
answer8



```
9. Probability that a normal random variable with mean 22 and variance 25  is greater than 29 

```{r}
#We use the same function for a value of 29
#and subtract from 1

1 - pnorm(29,22,sd=5)

```

10. If we wanted to know the probability for **less than** 17 instead of greater than:

```{r}

pnorm(17,22,sd=5)

```

11. And if we wanted to know the probability that it was less than 15 or greater than 25 (mutually exclusive possibilities):


```{r}
#The parentheses around the two parts aren't strictly necessary
#I just wanted to separate the two probabilities
(pnorm(15,22,sd=5)) +  (1-pnorm(25,22,sd=5))

```




