---
title: "Lab 3: July 15"
author: "POLS 3316, Summer 2022, University of Houston, Instructor: Tom Hanna"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## Part 1: Scalars, Vectors, Matrices and Arrays

### All of this is actually about *matrixes*, but scalars and vectors are the components of *matrices*, so we will get there last.

For now, a matrix is a collection of data in columns and rows.

### What is a scalar?

A scalar is a single number. A scalar is a matrix with dimension 1x1, 1 column by 1 row. A scalar is also a vector with only 1 element.

#### Scalars in R

To create a scalar in R, we can either assign a null value and add the value later, or we can assign a value when we creat it:

```{r}
a <- c()
cat('a is a null scalar ')
a

b <- 5
cat('b is a scalar with value 5')
b

a <- 1
cat('Now I have given a the value 1')
a
```



### What is a vector?

A vector is a list of numbers. It is either a single column or a single row. It is a matrix with dimension 1 x n . 

#### Vectors in R?

To create a vector in R, we can assign a series of values either by just listing them or through a variety of functions including operations on other vectors:

```{r}

C <- c(1,2,4,5,6,7,8,9,10)
cat('C is a vector ')
C

D <- C*b
cat('D is a vector created by multiplying b * C ')
D

```

```{r}

E <- seq(-5, 5, by = .5)
cat('E is a sequence of numbers from -5 to 5, by 0.5s ')
E

```

```{r}

F <- dnorm(E, mean = 2.5, sd = 0.5)
cat('F is a function of E and the normal distribution ')
F


```

### What is a matrix?

A *matrix* is a rectangular array or table of numbers, symbols, or expressions, arranged in rows and columns, used to represent some mathematical object or property. 

Practical examples of *matrices*? Spreadsheets and tables

#### Matrixes in R

In R, every *dataframe* or *table* or similar *object* is a matrix.



## Back to matrices in R

We built a scalar object and a vector object. How do we build a matrix object?

The most common way starting off is to build a dataframe. 

Let's use two of the vectors we built, C and D:

```{r}
data_cd <- data.frame(C,D)
View(data_cd)

```

Our dataframe doesn't have any row or column names. We can fix that:

```{r}

#First create a vector of the row names
cd_rownames <- c("Houston","Humble","Pasadena","League City",
                 "La Porte","Sugar Land","Pearland","Katy",
                 "Galveston")
#assign the vector using the rownames function        
rownames(data_cd) <- cd_rownames

#for the columns, we'll use the colnames function directly
#We can use either method with either function

colnames(data_cd) <- c("US dollars","Crypto assets")


```



## Things to know

**You don't need to know the details of multiplying matrixes, but at some point it will be useful to know that the sizes have to match.**

- Multiplying anything times a scalar is simply multiplying every element by that scalar. It's almost always feasible. 
- To do an operation on two vectors they must be the same length. (If they aren't, R may try to repeat the shorter matrix with results you may not expect.)
- When you multiply two matrixes, a row in one matrix is multiplied by a corresponding column from the second matrix. So, you can multiply a 2 (row) x 3 (column) matrix by a 3 (row) x 2 (column) matrix but you can't multiply a 2 x 3 by a 3 x 3. 

https://www.mathsisfun.com/algebra/matrix-multiplying.html

or 

**ask for help from someone with a good grasp of basic matrix algebra**

## Part 2: Measures of dispersion in R          

We're going back to the arrest data. We can use the R functions *var()* to get the **sample** *variance* and *sd()* to get the **sample** *standard deviation*. The numbers we figured were the **population** numbers. The difference in the formulas is that for the *variance*, the population version divides by the number of observations *n* and the sample version divides by *n-1*. We'll discuss the theoretical difference in detail this week, but for now the difference is this: In most cases we are only working with a sample of data and we are using it to estimate the full "population" or "universe" of data. The slight difference in formula is to account for this.  

For now, let's start by looking at the Arrest Data *sample variance* and *sample standard deviation.*


To get **sample variance** use the R function **var().**


```{r}
arrests_data <- USArrests[2:3]

urban_pop_variance <- var(arrests_data$UrbanPop)
cat('Urban Pop Percent Variance')
urban_pop_variance

assaults_variance <- var(arrests_data$Assault)
cat('Assaults Variance')
assaults_variance

mean(arrests_data$Assault)

```



To get **sample standard deviation** either take the **square root of sample variance** or use the R function **sd()**.

```{r}

cat('Standard deviations')
urban_pop_sd1 <- sd(arrests_data$UrbanPop)
urban_pop_sd1

urban_pop_sd2 <- sqrt(var(arrests_data$UrbanPop))
urban_pop_sd2

```


To get **population variance** multiply the R function var() times (n-1)/n:  var(x) * (n-1)/n

```{r}

urban_pop_popvar <- var(arrests_data$UrbanPop) * (50-1)/50
cat('Urban Pop population variance:')
urban_pop_popvar

assaults_popvar <- var(arrests_data$Assault) *(50-1)/50
cat('Assaults population variance:')
assaults_popvar

```


To get **population standard deviation**, the only option is to compute the population variance and takes its square root: **sqrt(populationvariance)** *You can't get it by multiplying the R function at least easily.*

```{r}

sqrt(urban_pop_popvar)
sqrt(assaults_popvar)

```





