---
title: 'Lab 8: July 22'
author: 'POLS 3316, Summer 2022, University of Houston, Instructor: Tom Hanna'
output: html_notebook
---

## Part I: Loops

Loops are funtions, or sets of instructions, which tell the computer to repeat an operation or set of operations several times. 
There are two kinds of loops in R, *for loops* and *while loops*. There is also a family of functions, the apply functions that act similary to loops with added advantages - *apply(), lapply(), tapply(), sapply(), vapply(),* and *mapply()*. Starting with understanding basic loops helps with understanding how the computer processes repeated instructions. It is also possible to nest loops within loops, which can be very powerful. Today we are going to cover basic simple, unnested *for loops* and *while loops*

#### For loops

A very simple loop tells R to start at some number, print it, then move to the next in the range:

```{r}
#i is an index that R uses to track where it is in the loop
#i increases by 1 on each pass until it reaches the end of the range
#1:5 is the range. The loop starts at i = 1 and runs until i = 5
#All this loop does is print the value of i
for (i in 1:5) {
  print(i)
}

```

You can also use a loop to perform vector operations. For example:

```{r}
#first, we need a vector to operate on

my_vector <- c(1,2,3,4,5)

#and a second vector to get the output
my_vector2 <- c()

for (i in 1:5) {
  my_vector2 <- c(my_vector2,my_vector * i)
}
print(my_vector2)

```

And we can loop through elements in a dataframe:

```{r}
#first we need some data
library(here) # load the here library
#create ojbect gotdata and have read.csv load the 
##file GOT_screentimes.csv in data subdirectory
gotdata <- read.csv(here("data","GOT_screentimes.csv")) 

for (name in gotdata) {
        Character <- (gotdata$name)
        Actor <- (gotdata$portrayed_by_name)
}
new_df <- data.frame(Actor,Character)
played_by <- " was played_by "

for (i in 1:191) {
      nth <- paste0(new_df$Character[i],played_by,new_df$Actor[i]) 
      print(nth)
}

```

###While loops

```{r}
#set the start point
i <- 1

#tell R to run the loop as long as is less than 29
while (i < 29) {
        #tell it what to do
        nth <- paste0(new_df$Character[i],played_by,
                      new_df$Actor[i]) 
      print(nth)
      #tell it to add 1 to i (otherwise, i never gets bigger and the loop runs forever)
      i = i + 1
}
```


## Part II: Frequency Distributions (Cont) and Correlations

